---
title: Algorithmic Notes on Face Recognition 
author: Oliver Dürr
output: pdf_document 
---
```{r, eval=TRUE,echo=TRUE}
  EVAL = TRUE
  ECHO = TRUE
```


## Introduction
These algorithmic notes eluciade the classifiers used in the SoE project *piVision*. For illustrating purpose the algrithms orignially implemented in python using the openCV and scikit library are partly reimplemented in R.

To evaluate the algorithm we used the `Rasberry Pi` with an attached picam to take several photographs for each person of the team in two batches, indoors and outdoors. We use the indoor pictures as a training set to learn the classifier and the outdoor batch for evaluation. (**TODO** or other way around?)

## What happend before (preprocessing)?
Before the feature extraction and classification we aligned and cropped the faces from the photos using the following method:

* **TODO** add the tricks with downscaling for performance increase. 
* In a first step faces are detected from the picture using the Viola-Jones Algorithm provided in `openCV` library. This boosting algorithm worked quite well and is descriped in XXX.


* **TODO** include a typical image.

* An ellipse around the face is set to zero (weak effect)

* The image is split into 3 vertcial stripes and each stripe is normalized (**TODO** provide details)

In the following the base-line classification algorithms are descriped and documented using the R. Slighty better results can be obtained when local binary pattern instead of the original grey values per pixels are used. This can be further enhanced by applying the LBP at different resolution on the image or parts of the image. See end to this report for some notes on these more adavanced techniques.

# Training data. 
The following loads the 226 aligned faces from the training set and plots several of them.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  trainingFile = "../../data/training_48x48_unaligned_large_expanded.p_R.csv.gz.csv"
  testFile = "../../data/testing_48x48_unaligned_large.p_R.csv.gz"
  #trainingFile = "../../data/training_48x48_unaligned_large.p_R.csv.gz"
  #testFile = "../../data/testing_48x48_unaligned_large.p_R.csv.gz" #TODO check this testfile can be wrong
  source("Utils.R")
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);
  X_training <- ddd[,-1]
  y_training <- ddd[,1]
  #perm <- sample(1:ncol(X_training), replace = FALSE)
  #X_training <- X_training[,perm]
  N <- sqrt(ncol(X_training))
  cat("Loaded Training set ", dim(X_training), " Dimension of pixels: ", N, "x", N)
  plotExamples(y_training,X_training, title = "Training ", mfrow=c(3,6))
```

# Test data
Same loading and plotting but now for the test-data.
```{r, eval=EVAL,echo=ECHO, fig.cap="Some examples of the test set.", fig.width=12, fig.height=6}
dumm <- read.table(testFile, sep=",", stringsAsFactors = FALSE)
ddd <- as.matrix(dumm);X_testing <- ddd[,-1];y_testing <- ddd[,1]
#X_testing <- X_testing[,perm]
N <- sqrt(ncol(X_testing))
cat("Loaded Test set ", dim(X_testing), " Dimension of pixels: ", N, "x", N, " number of y ", length(y_testing))
plotExamples(y_testing,X_testing, title = "Testing ", mfrow=c(3,6))
```

# Eigenfaces, the spucky images
We unroll the images into a $48 \cdot 48 = 2304$ dimensional vector\footnote{The data was in that format anyway}. Each image is thus a point in a 2304 dimensional vector space. We now want to find a new basis in the 2304 dimensional space so that the variance (of the pixel intensities) is maximal in the first basis component, second maximal in the second component (which also needs to be othogonal to the first one). This can be achieved by first writing each image row by row into a matrix of the dimension 226x2304. The correlation matrix can be obtained by scaling the matrix so that the mean columns is 0 and the sd is 1. Due to constant values in some columns (e.g. in the masked regions) we cannot force the standard deviation to be 1 and therefore just scale the mean. This corresponds to using the covariance instead of the correlation.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X = scale(X_training, center = TRUE, scale = FALSE)
  dim(X)
  max(colSums(X))
```
The covariance matrix is simply $X^T X$.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  cov = t(X)%*%X / dim(X)[1]
  dim(cov) #2304 x 2304
```
To get the direction of with maximal covariance, we simply have to calculate the Eigenvectors of the matrix
```{r, eval=FALSE,echo=ECHO, fig.width=12, fig.height=6}
  cov = t(X)%*%X / dim(X)[1]
  e <- eigen(cov)
```
We simply roll the 2304 dimensional Eigenvectors back into an 48x48 dimensional images and have the infamous  Eigenfaces. 

However, it take quite some time to calculate the Eigenvectors from a 2304 dimensional matrix. Here we can apply the following trick\footnote{There are also other numerical ways to calculate the principal components}. Note that the singular value decomposition (SVD) decomposes a Matrix $M$ under no assumptions! (**TODO: check** ) into:
$$ M = U \Sigma V^* $$ 
with U consisting of the Eigenvetors of $M M^*$ with $M^*$ being the conjugate transpose of $M$, $\Sigma$ is a diagonal matrix of the Eigenvalues and $V^*$ has the Eigenvectors of $M^* M$ in its columns. So to calculate the Eigenvalue of $X^T X$ we simply have to do an SVD of $X^T$ and take $U \in \mathcal{R}^{2304 x 226} $ which contains then the Eigenvectors of $X^T X$ as we wish. 
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
    res = svd(t(X) / sqrt(dim(X)[1]))
    dim(res$u)
    par(mfrow=c(3,6));par(mai=c(0.1,0.1,0.1,0.1))
    for (i in 1:18) {
      sm <- matrix(rev(res$u[,i]), ncol=N, byrow=TRUE)
      image(t(sm), useRaster=TRUE, main=NULL, col=gray.colors(255), axes = FALSE)
    }
```
Alternatively we can also do this in R using the `princomp` function.
**TODO** Explain the difference to R
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
fit <- princomp(t(X_training), cor=FALSE)
res.sc <- fit$scores # the principal components
par(mfrow=c(3,6));par(mai=c(0.1,0.1,0.1,0.1))
dim(res.sc)
for (i in 1:18) {
 # m <- scale(res.sc[,i])
  m <- res.sc[,i]
  sm <- matrix(rev(m), ncol=N, byrow=TRUE)
  image(t(sm), useRaster=TRUE, main=NULL, col=gray.colors(255), axes = FALSE)
}
par(mfrow=c(1,1))
```

## The labeled training in the Eigenspaces
In the following we plot the training and testdata in the space spanned by the first two Eigenvectors. We color the datapoint according to the person they belong to.

```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
par(mai=c(1.02,0.82,0.82,0.42))
pc.cr <- prcomp(X_training, center = FALSE)
X.train.pca <- pc.cr$x
dim(X.train.pca)
par(mfrow=c(1,2))
plot(X.train.pca[,1], X.train.pca[,2], col=y_training, main="Training")
dim(X.train.pca)
### ... and the test data with labels
par(mai=c(1.02,0.82,0.82,0.42))
X.test.pca <- predict(pc.cr, X_testing) #
dim(X.test.pca)
plot(X.test.pca[,1], X.test.pca[,2], col=y_testing, main="Testing")
par(mfrow=c(1,1))
```
We see some seperation of the individual persons already in the two dimensional slice. The classification algorithms below try to find a seperation of the space of `{r dim(pc.cr$x)` dimensions spanned by the Eigenvectors.\footnote{Sometime only the first few EV are used, but we did not find a benifit in this evaluation.} Note that the transformation is learned on the training set and then applied on the test set.

## Classification
We now use the PCA for dimension reduction and apply serveral algorithms for classification.

* Simple KNN in the space of the Eigenfaces which is what people call: "Face Regonition with Eigenfaces"
* Fisher LDA in the space of the Eigenfaces aka Fisherfaces
* SVM in the space of the Eigenfaces

The classifier is learned on the training set and the performance is evaluted to the test set. 

### Eigenfaces
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
library(class)
sum(knn(train = X.train.pca, test = X.test.pca, cl = y_training) == y_testing) / length(y_testing)  # 0.7348837
```
Result from our openCV-pipeline (Build 246) Eigenfaces 0.720930233

### Fisher Faces 
Here we use the Fisher LDA in which the the space has to smaller than the number of examples, hence we need to have a dimension reduction such as the PCA.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
library(MASS)
z <- lda(X.train.pca[,1:200], y_training)
res <- predict(z, X.test.pca[,1:200])
sum(res$class == y_testing) / length(y_testing) #0.9162791
```
Result from our openCV-pipeline (Build 246) Fisherfaces 0.893023256

### Running a SVM after PCA
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
require(e1071)
#table(as.factor(y_training))
model <- svm(X.train.pca, as.factor(y_training), kernel='linear', cost=1, scale = TRUE)
test.svm <- predict(model, X.test.pca)
#table(test.svm)
sum(test.svm == y_testing)/ length(y_testing) #0.9023 for all
```
Result from our openCV-pipeline (Build 246) 3|4 (Simple unrolling FE) 0.874418605. Properbly because scaling is note set. (**TODO**) Set scaling to true.

### Random forest on PCA
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
library("randomForest")
d.train <- data.frame(y = as.factor(y_training), X=X.train.pca)
model <- randomForest(y ~ ., data=d.train)
d.test <- data.frame(X=X.test.pca)
sum(predict(model, d.test) ==  y_testing)/ length(y_testing) 
sum(predict(model, d.test) ==  y_testing)
```
Not in our test cases.


# Mulinomial regression 
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
require(nnet)
d.train <- data.frame(y = as.factor(y_training), X=X.train.pca[,1:40])
model <- multinom(y ~ ., data=d.train)
d.test <- data.frame(X=X.test.pca[,1:40])
sum(predict(model, d.test) ==  y_testing)/ length(y_testing) 
```

#Notes on advanced algorithms and the performance of variantions
Note: the following methods are not implemented in R (yet).

The face detection can be enhanced by taking not the original pixels as a starting point but use methods to extract features from them. We tested the following feature extractors

  * 0 : Nothing needed for the original openCV faceregonizers
  * 1 : SimpleLBHPFeatureExtractor (Whole Image)
  * 2 : dito +1 Zooming
  * 3 : dito +2 Zooming
  * 4 : Unrolling FE, just use the pixels as in the above methods
  * 5 : LBPHFeatureExtractor, extract patches near the nose and eyes and calculate LBP at these regions in various zoomings. 

The extracted features are then used in the following classifiers

* LBHP: from openCV
* Fisher: FisherFaceRecognizer from openCV
* Eigen:  EigenfaceRecognizer from openCV
* SVN: SVM after PCA
* SVN1: Same as SVN but different implementation (small deviations are expected)

For the aligned 48x48 pictures from above we get the following results (rev = revision number of SVN repository)

rev  | LBPH	| Fisher | Eigen |	SVN/1 |	SVN/2 |	SVN/3	| SVN/4 | SVN/5 | SVN1/4| SVN1/4
-----|------|--------|-------|--------|-------|-------|-------|-------|-----|----
246  | 0.7813|	0.8930|	0.7209|	0.9720|	0.9674|	0.9488|	0.8744|	0.93488|	0.9813|	0.9116

We see that taking the LBHP features (feature extractor 1) instead of the original pixels (feature extractor 5) results in a performance increase of about 4% from `SVN/5` to `SVN/1`. 

The more complex feature extractors 2,3 and 5 play their strength only for larger images. The next table shows in the first line (rev 164) the result when using 120x120 images instead of the 48x48 ones. (Small remark images from Dejan just have been included starting from approx rev. <= 244).

rev  | LBPH   | Fisher | Eigen  |	SVN/1  |	SVN/2 |	SVN/3	 | SVN/4 | SVN/5 | SVN1/1	  | SVN1/4
-----|--------|--------|--------|--------|--------|--------|-------|-------|--------|----
164  | 0.8461 |0.9230  | 0.7252 |	0.9560 |	0.9505|	0.95604|0.89010| 0.9725|	0.9505|	0.901
163  | 0.2362 |0.2472  | 0.26373|	0.2472 |	0.2142|	0.14285|0.24725| 0.1703|	0.3076|	0.247

We see that for larger images the complex feature extractors become relevant, SVN/5 is now better than SVN/1.
However, the increase is not dramatic. The need for more complex FE should be evaluated probably with more challenging data-sets. We note also that for smaller images the simple FE 1 is as good as the complex FE 5 on larger images. This strongly suggest to use 48x48 images with the simple and LBHP feature extractor (**TODO** use in final revisions). Especially since the running time for the classification of one image of approx 250ms using  SVN/1 increases by a factor of approximately 4 **TODO loma check**. 

In revision 163 we switched off the alignment and observe a dramic drop of the performance pointing out the primordial importance of good alignment.


The effect of the different preprocessing method is evaluated in the following table:

rev  | LBPH   | Fisher | Eigen  |  SVN/1  |	SVN/2 |	SVN/3	 | SVN/4 | SVN/5 | SVN1/1	  | SVN1/4   | Comment
-----|--------|--------|--------|---------|-------|--------|-------|-------|--------|-------
161  |0.868	  |0.874   |0.522   |   0.868 |0.852  |0.841   |0.874  | 0.857 | 0.868  | 0.868 | Simple preprocessing
173  |0.863   |0.890   |0.692   |   0.956 |0.945  |0.945   |0.907  | 0.967 | 0.951  | 0.863 | Single hist. prep.
183  |0.830   |0.918   |0.725   |   0.956 |0.951  |0.956   |0.912  | 0.978 | 0.956  | 0.918 | No Masking of ellipse aroud face
164  |0.846   |0.9230  |0.7252  |   0.956 |0.951  |0.956   |0.890  | 0.972 | 0.9505 |	0.901 | Full preprocessinfg


### Full preprocessing
The best preprocessing so far splits the image into 3 vertical parts and equalizers the histogram for each part of the  seperately
```
  X1 = cv2.equalizeHist(X[0:, 0:SIZE_FOR_PCA[1] / 3])
  X2 = cv2.equalizeHist(X[0:, SIZE_FOR_PCA[1] / 3:(SIZE_FOR_PCA[1] * 2 / 3)])
  X3 = cv2.equalizeHist(X[0:, (SIZE_FOR_PCA[1] * 2 / 3):SIZE_FOR_PCA[1]])
  X = np.append(np.append(X1, X2, 1), X3, 1)
```

### Simple preprocessing
Use histogram equalization
```
 X = cv2.equalizeHist(X)
```

### Simple preprocessing
Is actually not to simple
```
  # From https://github.com/bytefish/facerec/blob/master/py/facerec/preprocessing.py
  alpha = 0.1
  tau = 10.0
  gamma = 0.2
  sigma0 = 1.0
  sigma1 = 2.0
  
  sigma0 = 2.0
  sigma1 = 5.0
  
  X = np.array(X, dtype=np.float32)
  X = np.power(X, gamma)
  X = np.asarray(ndimage.gaussian_filter(X, sigma1) - ndimage.gaussian_filter(X, sigma0))
  X = X / np.power(np.mean(np.power(np.abs(X), alpha)), 1.0 / alpha)
  X = X / np.power(np.mean(np.power(np.minimum(np.abs(X), tau), alpha)), 1.0 / alpha)
  X = tau * np.tanh(X / tau)
```

## Effect of the  alignment.
Since the effect of the alignment is so important, we like to elucidate a bit on it in the following.

Below is a comparison of the pictures without and with the simple "two eyes" same position alignment. 

### Unaligned images
The following images are not aligned, the cropping happend using the detected face (Viola-Jones).
```{r, eval=EVAL,echo=FALSE, fig.width=12, fig.height=6}
  source("Utils.R")
  trainingFile = "../../data/training_48x48_unaligned_large.p_R.csv.gz"
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);X_training <- ddd[,-1];y_training <- ddd[,1]
  plotExamples(y_training,X_training, title = "Training ", mfrow=c(3,6))
```

### Aligned images
```{r, eval=EVAL,echo=FALSE, fig.width=12, fig.height=6}
  trainingFile = "../../data/training_48x48_aligned_large.p_R.csv.gz"
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);X_training <- ddd[,-1];y_training <- ddd[,1]
  plotExamples(y_training,X_training, title = "Training ", mfrow=c(3,6))
```
It's really hard to believe that the small distortions have such a strong effect. However the algorithms are not very robust against little transformations. Accually the Algorithms don't care if you reshuffel the image, the performance is as good as you let it run on those images.


### Aligned images (reshuffeled)
```{r, eval=EVAL,echo=FALSE, fig.width=12, fig.height=6}
  trainingFile = "../../data/training_48x48_aligned_large.p_R.csv.gz"
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);X_training <- ddd[,-1];y_training <- ddd[,1]
  perm <- sample(1:ncol(X_training), replace = FALSE)
  X_training <- X_training[,perm]
  #X_training <- X_training[,perm]
  plotExamples(y_training,X_training, title = "T. (reshuffeld)", mfrow=c(3,6))
```
Do you see that the first three pictures belong to the same person, well the algorithm does. I really did the experiment the performance stays the same, as long as you perform the same reshuffeling on each image!

.           |Eigenfaces | Fisher | SVM
------------|-----------|--------|-----
Reshuffeled | 0.7349    | 0.9163 | 0.9023
Original    | 0.7349    | 0.9163 | 0.9023

For the more advaced methods (not implemented in R see below) this is not true anymore but still the performance decreases significantly if you don't align.



## Other methods 
We further evaluated the preformance on a convolutional neural network. We used aligned and unaligned faces of size 48x48 and downscaled them to 28x28 pixels. The training set stayed the same but the testset had to be split into a validation and testset. We used the theano sample code (http://www.deeplearning.net/tutorial/lenet.html#lenet). We obtained an error of 16% for the aligend faces with a learning rate of 0.001. Note that the learning rate had to be manually altered. For the unaligned faces the error had be 50% at the same learning rate. We also tried whitening on the whole image. However this resulted in a error of 30% inferior to not whitening.




### Evaluation with existing 
http://vis-www.cs.umass.edu/lfw/











