---
title: Covariance, SVD, PCA and Whitening
author: Oliver Dürr
output: pdf_document 
---
```{r, eval=TRUE,echo=FALSE}
  EVAL = TRUE
  ECHO = TRUE
```


## Introduction
These notes are about PCA, SVD and whitening. They are somehow motivated by the UDFL tutorial on whitening (http://ufldl.stanford.edu/wiki/index.php/Implementing_PCA/Whitening).

We start with a simple toy example to show what happens in a SVD. First we consider the situation where we have 30 examples each having 4 pixels. We code this (in contrast to the piVision project) as a 4x30 matrix (columns code examples). 

### Calculation of the covariance matrix
We want to investigate the covariance between the pixels. And show that such a matrix can be calculated as $\tilde{X} \tilde{X'}$ with $\tilde{X}$ is $X$ normalized so that the pixels have zero mean.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  # http://www.mathworks.ch/help/matlab/ref/svd.html
  ncol = 30  
  X.org   = matrix(rnorm(4 * ncol), ncol = ncol)
  colnames(X.org) = paste("Example", 1:ncol)
  rownames(X.org) = paste("Pixel", 1:4)
  avg = rowMeans(X.org)
  X = X.org - avg
  cov.r = cov(t(X.org))
  XXT = (X %*% t(X)) / (ncol - 1)   
  max(abs(XXT - cov(t(X)))) #This confirms that the covariance of the pixels is XX'
```

### Rotate into new pixel space
Next we want to rotate into the new pixel space, in which the covariance matrix is diagonal. This can be done via $\vec{x}_{rot} = U' \vec{x}$ where $\vec{x}$ is a 4 dimensional pixel vector.

```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  e = eigen(XXT)
  U = e$vectors
  x_rot = t(U) %*% X[,1]     # Rotation for one pixel vector
  X_rot = t(U) %*% X         # Rotation for all 30 vectors
  plot(X_rot[1,], X_rot[2,]) # Plotting the first 2 new pixel
  cov(X_rot[1,], X_rot[2,])  # Covariance is diagonal in the new space
  cov(t(X_rot))              # Covariance is diagonal in the new space
```

### Comparison with R
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  fit <- princomp(t(X.org), cor=FALSE)
  res.sc <- fit$scores # the principal components
  max(abs(t(res.sc) - X_rot))
```

### Using svd instead of Eigenvalues $X$ of $X X'$ don't care
This is really strange, one can use SVD of X or XXT.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  XXT = (X %*% t(X)) / (ncol - 1) #Constant terms does not matter
  eigen(XXT)$vectors
  svd(X)$u
  svd(XXT)$u
  svd(t(X))$v
```


### Whitening: make covariance in new pixel space the identity
Since the covariance matrix given by $X_{rot} X'_{rot}$ is diagonal with the diagonal given by the Eigenvalues we can make it the identity by dividing $X_{rot}$ with $\sqrt{\lambda_i}$. 
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X_rot_w = X_rot / sqrt(e$values)
  cov(t(X_rot_w))
```
If we now go back to the original space, we find that the covariance matrix is diagonal.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X_w = U %*% X_rot_w
  X_w_1 = U %*% diag(1/sqrt(e$values)) %*% t(U) %*% X 
  cov(t(X_w_1))
```

### In conclusion: 
If X is coded so that the pixels are the rows, then the transformation
$$
  X_w = U * diag(1/\sqrt{\lambda_i}) * U' * X
$$
removes the correlation between the pixels it *whitens* or *spheres* the image. To be save against small Eigenvectors a small constant typically $10^-5$ is added to the Eigenvalues $\lambda_i$.

### More pixels then examples
If we have more pixels then examples, we simply transpose.

```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  ncol = 3
  X.org   = matrix(rnorm(4 * ncol), ncol = ncol)
  colnames(X.org) = paste("Example", 1:ncol)
  rownames(X.org) = paste("Pixel", 1:4)
  
  # Transpose
  X.org = t(X.org)
  avg = rowMeans(X.org)
  
  X = X.org - avg
  res = svd(X)
  U = res$u
  e = res$d
  
  es = eigen(X %*% t(X))
  U = es$vectors
  e = es$values 

  X_w_t = U %*% diag(1/sqrt(e + 1e-5)) %*% t(U) %*% X

  # Transpose back
  X_w = t(X_w_t)
  cov(X_w)
```

# Training data. 
The following loads the 226 aligned faces from the training set and plots several of them.
```{r, eval=TRUE, echo=FALSE, fig.width=12, fig.height=6}
  source("Utils.R")
  trainingFile = "../../data/training_48x48_aligned_large.p_R.csv.gz"
  testFile = "../../data/testing_48x48_aligned_large.p_R.csv.gz"
  #trainingFile = "../../data/training_48x48_unaligned_large.p_R.csv.gz"
  #testFile = "../../data/testing_48x48_unaligned_large.p_R.csv.gz" #TODO check this testfile can be wrong
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);
  X_training <- ddd[,-1]
  y_training <- ddd[,1]
  #perm <- sample(1:ncol(X_training), replace = FALSE)
  #X_training <- X_training[,perm]
  N <- sqrt(ncol(X_training))
  cat("Loaded Training set ", dim(X_training), " Dimension of pixels: ", N, "x", N)
  plotExamples(y_training,X_training, title = "Training ", mfrow=c(3,6))
```

# Test data
Same loading and plotting but now for the test-data.
```{r, eval=TRUE,echo=FALSE, fig.cap="Some examples of the test set.", fig.width=12, fig.height=6}
dumm <- read.table(testFile, sep=",", stringsAsFactors = FALSE)
ddd <- as.matrix(dumm);X_testing <- ddd[,-1];y_testing <- ddd[,1]
#X_testing <- X_testing[,perm]
N <- sqrt(ncol(X_testing))
cat("Loaded Test set ", dim(X_testing), " Dimension of pixels: ", N, "x", N, " number of y ", length(y_testing))
plotExamples(y_testing,X_testing, title = "Testing ", mfrow=c(3,6))
```


## Using pictures (Training set)
```{r, eval=FALSE,echo=ECHO, fig.width=12, fig.height=6}
  X.org = t(X_training)
  
  dim(X.org)
  avg = rowMeans(X.org)
  X = X.org - avg  
  XXT = (X %*% t(X)) / (ncol(X) - 1)
  dim(XXT)  
  e = eigen(XXT)
  U = e$vectors
  X_w_1 = U %*% diag(1/sqrt(e$values + 1)) %*% t(U) %*% X
  X_w_1 = U %*% diag(c(1/sqrt(e$values[1:100]), rep(0, length(e$values) - 100))) %*% t(U) %*% X
  #dumm = U %*% diag(1/sqrt(e$values + 0.001)) %*% t(U) 
  #Same as self.components_ in the python cod
  dim(X_w_1)

  X_w = t(X_w_1)
  dim(X_w)
  plotExamples(y_training,X_w, title = "Training whitend", mfrow=c(3,6))

  cov(X_w[,10], X_w[,10]) #1 Pixel 1 and Pixel 1 
  cov(X_w[,20], X_w[,20]) #1 Pixel 2 and Pixel 2
  cov(X_w[,21], X_w[,22]) #1e-13 Pixel 1 and 2
```




