---
title: Covariance, SVD, PCA and Whitening
author: Oliver Dürr
output: pdf_document 
---
```{r, eval=TRUE,echo=FALSE}
  EVAL = TRUE
  ECHO = TRUE
```


## Introduction
These algorithmic notes are about PCA, SVD and whitening. They are somehow motivated by the UDFL tutorial on whitening (http://ufldl.stanford.edu/wiki/index.php/Implementing_PCA/Whitening).

We start with a simple toy example to show what happens in a SVD. First we consider the situation where we have 30 examples each having 4 pixels. We code this (in contrast to the piVision project) as a 4x30 matrix (columns code examples). Note that this is in the reversed order compared to the coding we apply in the piVision project, but it's more convenient to think in column vectors.

### Calculation of the covariance matrix
We want to investigate the covariance between the pixels. And show that such a matrix can be calculated as $\tilde{X} \tilde{X'}$ with $\tilde{X}$ is $X$ normalized so that the pixels have zero mean.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  # http://www.mathworks.ch/help/matlab/ref/svd.html
  ncol = 30  
  X.org   = matrix(rnorm(4 * ncol), ncol = ncol)
  colnames(X.org) = paste("Example", 1:ncol)
  rownames(X.org) = paste("Pixel", 1:4)
  avg = rowMeans(X.org)
  X = X.org - avg
  cov.r = cov(t(X.org))
  XXT = (X %*% t(X)) / (ncol - 1)   
  max(abs(XXT - cov(t(X)))) #This confirms that the covariance of the pixels is XX'
```

### Rotate into new pixel space
Next we want to rotate into the new pixel space. This can be done via $\vec{x}_{rot} = U' \vec{x}$ where $\vec{x}$ is a 4 dimensional pixel vector.

```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  e = eigen(XXT)
  U = e$vectors
  x_rot = t(U) %*% X[,1]     # Rotation for one pixel vector
  X_rot = t(U) %*% X         # Rotation for all 30 vectors
  plot(X_rot[1,], X_rot[2,]) # Plotting the first 2 new pixel
  cov(X_rot[1,], X_rot[2,])  # Covariance is diagonal in the new space
  cov(t(X_rot))              # Covariance is diagonal in the new space
```

### Comparison with R
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  fit <- princomp(t(X.org), cor=FALSE)
  res.sc <- fit$scores # the principal components
  max(abs(t(res.sc) - X_rot))
```

### Using svd instead of Eigenvalues $X$ of $X X'$ don't care
This is really strange, one can use SVD of X or XXT.
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X.org   = matrix(rnorm(4 * ncol), ncol = ncol)
  avg = rowMeans(X.org)
  X = X.org - avg
  XXT = (X %*% t(X)) / (ncol - 1) #Constant terms does not matter
  eigen(XXT)$vectors
  svd(X)$u
  svd(XXT)$u
  svd(t(X))$v
```


## Whitening: make covariance in new pixel space the identity
Since the covariance matrix is given by $X_rot_w$

```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X_rot_w = X_rot / sqrt(e$values)
  cov(t(X_rot_w))
```

### Go back to original space 
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X_w = U %*% X_rot_w
  X_w_1 = U %*% diag(1/sqrt(e$values)) %*% t(U) %*% X
  cov(t(X_w_1))
```

## More pixels then examples
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  ncol = 3
  X.org   = matrix(rnorm(4 * ncol), ncol = ncol)
  colnames(X.org) = paste("Example", 1:ncol)
  rownames(X.org) = paste("Pixel", 1:4)
  avg = rowMeans(X.org)
  X = X.org - avg
  XXT = (X %*% t(X)) / (ncol - 1)   
  e = eigen(XXT)
  U = e$vectors
  X_w_1 = U %*% diag(1/sqrt(e$values + 1e-5)) %*% t(U) %*% X
  cov(t(X_w_1))
  
  # Other possibility
  X.org = t(X.org)
  ncol = ncol(X.org)
  nrow = nrow(X.org)
  colnames(X.org) = paste("Example", 1:ncol)
  rownames(X.org) = paste("Pixel", 1:3)
  avg = rowMeans(X.org)
  X = X.org - avg
  XXT = (X %*% t(X)) / (ncol - 1)   
  e = eigen(XXT)
  U = e$vectors
  X_w_1 = U %*% diag(1/sqrt(e$values + 1e-5)) %*% t(U) %*% X
  t(X_w_1)
  cov(t(X_w_1))
```




# Training data. 
The following loads the 226 aligned faces from the training set and plots several of them.
```{r, eval=TRUE, echo=FALSE, fig.width=12, fig.height=6}
  source("Utils.R")
  trainingFile = "../../data/training_48x48_aligned_large.p_R.csv.gz"
  testFile = "../../data/testing_48x48_aligned_large.p_R.csv.gz"
  #trainingFile = "../../data/training_48x48_unaligned_large.p_R.csv.gz"
  #testFile = "../../data/testing_48x48_unaligned_large.p_R.csv.gz" #TODO check this testfile can be wrong
  dumm <- read.table(trainingFile, sep=",", stringsAsFactors = FALSE)
  ddd <- as.matrix(dumm);
  X_training <- ddd[,-1]
  y_training <- ddd[,1]
  #perm <- sample(1:ncol(X_training), replace = FALSE)
  #X_training <- X_training[,perm]
  N <- sqrt(ncol(X_training))
  cat("Loaded Training set ", dim(X_training), " Dimension of pixels: ", N, "x", N)
  plotExamples(y_training,X_training, title = "Training ", mfrow=c(3,6))
```

# Test data
Same loading and plotting but now for the test-data.
```{r, eval=TRUE,echo=FALSE, fig.cap="Some examples of the test set.", fig.width=12, fig.height=6}
dumm <- read.table(testFile, sep=",", stringsAsFactors = FALSE)
ddd <- as.matrix(dumm);X_testing <- ddd[,-1];y_testing <- ddd[,1]
#X_testing <- X_testing[,perm]
N <- sqrt(ncol(X_testing))
cat("Loaded Test set ", dim(X_testing), " Dimension of pixels: ", N, "x", N, " number of y ", length(y_testing))
plotExamples(y_testing,X_testing, title = "Testing ", mfrow=c(3,6))
```




## Using pictures (Training set)
```{r, eval=EVAL,echo=ECHO, fig.width=12, fig.height=6}
  X.org = X_training
  avg = rowMeans(X.org)
  X = X.org - avg
  ncol = dim(X)[1]
  XXT = (X %*% t(X)) / (ncol - 1)   
  e = eigen(XXT)
  U = e$vectors
  X_w_1 = U %*% diag(1/sqrt(e$values + 1e-5)) %*% t(U) %*% X
  plotExamples(y_training,X_w_1, title = "Training whitend", mfrow=c(3,6))
  cov(t(X_w_1))[1:10,1:10]
```




